Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-09 ind_files/01_Introduction.tex
Add some references for importance of first principles approach

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size, thus restricting their validity in a world where population sizes routinely fluctuate. This is unfortunate, since empirical studies indicate that populations with fluctuating population size can exhibit evolutionary dynamics that are starkly different from the predictions of constant population models~\citep{sanchez_feedback_2013,chavhan_larger_2019,chavhan_interplay_2021}. Theoretical studies concerned with evolution in populations of non-constant size usually impose deterministic and typically phenomenological rules for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. These rules further usually do not depend on population composition. Such models are thus somewhat artificial since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death~\citep{metcalf_why_2007,doebeli_towards_2017}. Notably, the most general framework we have, the Price equation, is typically formulated in a deterministic setting ~\citep{page_unifying_2002,frank_natural_2012,queller_fundamental_2017,lion_theoretical_2018,day_price_2020} that ignores stochasticity (but see~\citet{rice_stochastic_2008} and~\citet{rice_universal_2020} for a discrete time formulation that is stochastic but, as far as I can tell, is dynamically insufficient, just like the original formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real systems of interest.

>>>>

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size, thus restricting their validity in a world where population sizes routinely fluctuate~\citep{lambert_population_2010}. This is unfortunate, since empirical studies indicate that populations with fluctuating population size can exhibit evolutionary dynamics that are starkly different from the predictions of constant population models~\citep{sanchez_feedback_2013,chavhan_larger_2019,chavhan_interplay_2021}. Theoretical studies concerned with evolution in populations of non-constant size usually impose deterministic and typically phenomenological rules for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. These rules further usually do not depend on population composition. Such models are thus somewhat artificial since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death~\citep{metcalf_why_2007,lambert_population_2010, geritz_mathematical_2012,doebeli_towards_2017}. Notably, the most general framework we have, the Price equation, is typically formulated in a deterministic setting ~\citep{page_unifying_2002,frank_natural_2012,queller_fundamental_2017,lion_theoretical_2018,day_price_2020} that ignores stochasticity (but see~\citet{rice_stochastic_2008} and~\citet{rice_universal_2020} for a discrete time formulation that is stochastic but, as far as I can tell, is dynamically insufficient, just like the original formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real systems of interest.

### 2023-04-09 ind_files/02_math_background.tex
Add some references for importance of first principles approach

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. As I explain in Appendix \ref{App_SDE_FPE}, the Fokker-Planck equation can be viewed as a `conservation law' for probability that mathematically expresses the common-sense observation that the sum (or integral) of probabilities over the entire state space cannot change over time (since it must always equal 1). If the function $G$ is independent of $x$, it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve).

@@

\subsection{Density-dependence and the intuition for system-size expansions via ecology}\label{sec_intuition_sys_size}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations of interest to population dynamics, at an individual level, births and deaths of individuals are affected by ecological rules that depend on the local population density and not directly on the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ through the introduction of a `system size parameter' $K$. In physics and chemistry, $K$ is usually the total volume of a container in which physical or chemical reactions take place and is thus a `hard' limit on the number of discrete values allowed. In ecology, we will only impose a `soft' limit by requiring that births and deaths must scale with $K$ in a way that the population almost surely cannot grow indefinitely if $K$ is finite, reflecting the empirical fact that the total amount of resources in the world is limited (of course, this should also be reflected in how birth and death rates vary as functions of population density). Thus, in our systems, $K$ will manifest as some fundamental limit on resources, such as habitat size or carrying capacity. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms\footnote{If this sounds handwavy to you, see chapter 11, section 3 in~\citet{ethier_markov_1986} for a more rigorous treatment.}. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'~\citep{ewens_mathematical_2004} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter.

>>>>

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004,barton_mathematical_2019} and pure mathematics~\citep{oksendal_stochastic_1998} literature. As I explain in Appendix \ref{App_SDE_FPE}, the Fokker-Planck equation can be viewed as a `conservation law' for probability that mathematically expresses the common-sense observation that the sum (or integral) of probabilities over the entire state space cannot change over time (since it must always equal 1). If the function $G$ is independent of $x$, it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve).

@@

\subsection{Density-dependence and the intuition for system-size expansions via ecology}\label{sec_intuition_sys_size}
The fundamental idea behind the system-size expansion relates to the nature of the jumps between successive states of a birth-death process. In most situations of interest to population dynamics, at an individual level, births and deaths of individuals are affected by ecological rules that depend on the local population density and not directly on the total population size. Despite this, the jumps themselves occur in terms of the addition (birth) or removal (death) of a \emph{single individual} from the population. If there are many individuals, each individual contributes a negligible amount to the density, and thus, the discontinuous jumps due to individual-level births or deaths can look like a small, \emph{continuous} change in population density. This is the essential idea behind the system-size expansion. The name derives from the formalization of this idea as a change of variable from the discrete values $\{0,1,2,\ldots,n-1,n,n+1,\ldots\}$ to the approximately continuous values $\{0,1/K,2/K,\dots,x-1/K,x,x+1/K,\ldots\}$ through the introduction of a `system size parameter' $K$. In physics and chemistry, $K$ is usually the total volume of a container in which physical or chemical reactions take place and is thus a `hard' limit on the number of discrete values allowed. In ecology, we will only impose a `soft' limit by requiring that births and deaths must scale with $K$ in a way that the population almost surely cannot grow indefinitely if $K$ is finite, reflecting the empirical fact that the total amount of resources in the world is limited (of course, this should also be reflected in how birth and death rates vary as functions of population density). Thus, in our systems, $K$ will manifest as some fundamental limit on resources, such as habitat size or carrying capacity. When $K$ is large, the fact that transitions occur in units of a small value $1/K$ can be exploited via a Taylor expansion of the transition rates in the Master equation, which then yields a Fokker-Planck equation upon neglecting higher order terms\footnote{If this sounds handwavy to you, see chapter 11, section 3 in~\citet{ethier_markov_1986} for a more rigorous treatment.}. A similar approximation is well-known (ever since Fisher) in theoretical population genetics, where it goes by the name of the `diffusion approximation'~\citep{ewens_mathematical_2004,barton_mathematical_2019} or `continuum limit'~\citep{czuppon_understanding_2021}, and has been heavily used by Kimura~\citep{crow_introduction_1970} in his stochastic models. However, the population genetics version of the approximation usually either relies on total population size being fixed~\citep{crow_introduction_1970, lande_natural_1976,ewens_mathematical_2004} or is conducted in an ad-hoc manner without specifying an explicit system size parameter.

### 2023-04-09 ind_files/06_nontechnical_discussion.tex
Add some references for importance of first principles approach

Stochastic models in biology often exhibit behaviors that are markedly different from their deterministic limits~\citep{jafarpour_noise-induced_2017,boettiger_noise_2018,jhawar_noise-induced_2020,coomer_noise_2022}. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems~\citep{black_stochastic_2012,hastings_transients_2004,shoemaker_integrating_2020,schreiber_does_2022}. In the context of population biology, the finiteness of populations is an important source of stochasticity that can lead to behavior not captured in corresponding infinite population limits~\citep{black_stochastic_2012,rogers_demographic_2012,debarre_evolutionary_2016,delong_stochasticity_2023}. When actually incorporating this stochasticity into eco-evolutionary theory, it is important to build up the stochastic theory from first principles instead of adding noise to existing deterministic models in an ad-hoc fashion, since the latter procedure can very easily lead to inconsistent models~\citep{strang_how_2019}. Indeed, several theorists have called for a reformulation of eco-evolutionary dynamics of finite populations starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic~\citep{metcalf_why_2007,geritz_mathematical_2012,doebeli_towards_2017}.

>>>>

Stochastic models in biology often exhibit behaviors that are markedly different from their deterministic limits~\citep{jafarpour_noise-induced_2017,boettiger_noise_2018,jhawar_noise-induced_2020,coomer_noise_2022}. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems~\citep{black_stochastic_2012,hastings_transients_2004,shoemaker_integrating_2020,schreiber_does_2022}. In the context of population biology, the finiteness of populations is an important source of stochasticity that can lead to behavior not captured in corresponding infinite population limits~\citep{black_stochastic_2012,rogers_demographic_2012,debarre_evolutionary_2016,delong_stochasticity_2023}. When actually incorporating this stochasticity into eco-evolutionary theory, it is important to build up the stochastic theory from first principles instead of adding noise to existing deterministic models in an ad-hoc fashion, since the latter procedure can very easily lead to inconsistent models~\citep{strang_how_2019}. Indeed, several theorists have called for a `bottom-up' reformulation of eco-evolutionary dynamics from the first principles of stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic~\citep{metcalf_why_2007,lambert_population_2010, geritz_mathematical_2012,doebeli_towards_2017}.
